Topological Eulerian Synthesis of Slow Motion Periodic Videos

### 2018-02-05 Paper/paper.tex
another small change
pass on paper

Our approach to slow motion templates is {\em Eulerian}; that is, we process the video pixel by pixel with no tracking.  Eulerian approaches for video synthesis are attractive due to their simplicity and ease of implementation.  For instance, a simple Fourier bandpass filtering and amplifying pixels can successfully elucidate subtle periodic motions in videos \cite{wu2012eulerian, wadhwa2013phase}.  Additionally, time-domain Eulerian approaches have been used to synthesize infinitely playing ``video textures''~\cite{schodl2000video}, as well as looping video templates for more general scenarios~\cite{Liao2013VideoLoops,Liao2015VideoLoops}.  There are some pitfalls with Eulerian techniques, however.  They are known to fail for large motions, and also require the user to manually specify frequencies of interest \cite{wu2012eulerian, wadhwa2013phase}.  Furthermore, period-to-period drift that is due to a moving camera or small variation in motion can be problematic, as well as unrelated motion in the background \cite{stauffer1999adaptive}.

@@

% MSB: opportunity to slim paper by cutting examples 
To address the problems with an Eulerian representation, we devise a geometric/topological Eulerian framework by constructing {\em sliding window embeddings} of videos (Section~\ref{sec:slidingwindow}).  Sliding window embeddings, or delay reconstructions, have found a diverse array of applications in activity recognition \cite{frank2010activity,venkataraman2016shape}, video analysis \cite{tralie2017quasi}, and motion capture analysis \cite{venkataraman2016shape}.  Sliding window embeddings of periodic time series, in particular, form samples of a topological loop \cite{perea2015sliding}, as do generalized multivariate sliding windows on periodic video data, regardless of the type of motion present \cite{traliehigh, tralie2017quasi}.  Furthermore, the sliding window embedding of an Eulerian video provides a form of regularization that mitigates the effects of drift and leads to a cleaner loop~\cite{tralie2017quasi}.  Our geometric framework also allows us to use topological data analysis (TDA) \cite{edelsbrunner2000topological,edelsbrunner2008persistent,edelsbrunner2010computational,carlsson2009topology,ghrist2014elementary} and fundamental frequency estimation \cite{Mcleod05asmarter} to autotune the spatial and temporal scales of the periodic motion, respectively, so that no user intervention is required.  We use Laplacian Eigenmaps~\cite{belkin2003laplacian} computed on the sliding window embedding to obtain a circular phase $\phi \in [0, 2\pi]$ for each window.  We then reorder the embedding using $\phi$, and reorder the video by having windows that overlap in time vote on final pixels in each frame of the template, helping to further mitigate drift (Section~\ref{sec:cyclereordering}).  As an illustration, Figure~\ref{fig:ConceptFigure} shows an example of our technique on a sampled 1D periodic time series.  There are only 12 samples per period in the original signal, so the details of each period are coarse and noisy.  However, once resorted, we get a fine-detailed representation of one period.

% MSB: opportunity to slim paper by cutting examples 
Many prior works use a combination of sliding window embeddings and topological data analysis, detecting chatter in mechanical systems \cite{khasawneh2016chatter}, and quantifying periodic activites in motion capture data \cite{vejdemo2015cohomological, venkataraman2016persistent}. However, our approach in using a sliding window jointly with TDA to {\em synthesize} a new result is novel. Related works have used the graph Laplacian to rearrange images around a loop as a pre-processing step for structure from motion \cite{averbuch2015ringit}, and to reorder the frames of microscope images from a developing embryo around a loop using vector diffusion maps~\cite{dsilva2015diffusionvecordering}.  However, our use of the graph Laplacian to instead {\em parameterize} time series is unique. One can also view the combination of TDA and Laplacian eigenmaps as a simpler alternative to cohomology circular coordinates \cite{de2011persistent,vejdemo2015cohomological} for the special case of periodic data.
%Our approach is also the opposite of the approach in \cite{bendich2011improving}, which performs diffusion maps before TDA. - MSB: this is related work, not opposite work :)

@@

Figure~\ref{fig:ConceptFigure} shows the different parts of our method for a 1D periodic signal for illustration purposes. In general, the actual input to our method is a 3-channel video of $N$ frames, each frame of resolution $W \times H$. We treat each frame as a point in Euclidean space of dimension $\mathbb{R}^{3 \cdot W \cdot H}$, and denote the set of all videos by the sequence $\mb{X}(t) \in \mathbb{R}^{(3 \cdot W \cdot H)}$ for integer-valued $t \in [1,N]$. The steps of our approach outlined in Figure~\ref{fig:ConceptFigure}, which we split into two parts: an analysis part (a), which culminates in phase estimates $\phi$, and a synthesis part (b), in which the template is created based on information in (a).  For the analysis part, we do not rely on the {\em extrinsic} geometry of the set of points $\mb{X}(t)$, only its set of {\em pairwise distances}. Thus, we can project all of the points into a lower-dimensional space and apply the set of techniques to the projection, assuming that pairwise distances are preserved. Since $N \ll (3 \cdot W \cdot H)$, the frames can be projected into a $N-1$-dimensional space such that all distances are preserved. Denoting this projection by $\mb{P} \in \mathbb{R}^{(N-1) \times (3 \cdot W \cdot H)}$, our method for the steps in (a) operates on $\mb{Z}(t) = \mb{P} \mb{X}(t)$.  We also use the third layer of a Gaussian pyramid on each frame in place of the raw frames before applying $\mb{P}$ to mitigate drift.

@@

The sliding window video embedding \cite{cao1998dynamics,traliehigh,tralie2017quasi} for $\mb{Z}$ is defined as
\begin{equation}
	SW_{d}[\mb{Z}(t)] = \left[ \begin{array}{c} \mb{Z}(t) \\ \mb{Z}(t + 1) \\ \vdots \\ \mb{Z}(t + (d-1))  \end{array} \right] \in \mathbb{R}^{(N-1)\cdot(d)},
\end{equation}
where $d$ is the dimension of the embedding. As shown by Takens~\cite{takens1981detecting}, a sliding window of dimension $2m+1$ of even a single generic observation function of a dynamical system of intrinsic dimension $m$ is sufficient to reconstruct a topological embedding of the underlying trajectory in the original state space.  For periodic signals, this state space is a torus, and $d$ should be twice the number of harmonics present to injectively reconstruct loops on that torus \cite{perea2015sliding}.  The same is true of videos in general \cite{tralie2017quasi}.  Furthermore, the sliding window length $(d-1)$ maximizes the roundness of the embeddings when $(d-1) = k T$ for some integer $k$, where $T$ is the fundamental period.  Since we will use topological data analysis (TDA) to find the scale of the loop, and since TDA measures roundness, it is important that we tune our window size to be an integer multiple of the period.  To do this, we perform 1D ISOMAP \cite{tenenbaum2000global} on the original frames of the video to generate a 1D surrogate signal, and we perform an autocorrelation-based fundamental frequency estimation \cite{Mcleod05asmarter} on this signal to estimate the period and, hence, the appropriate $d$.  This estimation often returns integer multiples of the fundamental period, which is fine for our purposes.

@@

We now examine the unweighted Laplacian of an ideal model to motivate our approach.  Suppose we have extracted a sliding window embedding $SW_{d}$ for a video $\mb{Z}$ which repeats itself exactly every $T$ frames $k$ times total, for a total of $N = kT$ windows.  Let us further suppose it is possible to choose an $\sigma$ so that the unweighted adjacency matrix contains an entry of 1 for edges corresponding to windows that are adjacent in time, as well as for the corresponding windows at, before, and after repetitions of the window at intervals of $T$; that is, the Laplacian is a symmetric, {\em circulant} matrix defined as

@@

The smallest two nonzero eigenvalues occur when $m = k$, corresponding to the eigenvectors \footnote{This generalizes the circle graph used in \cite{averbuch2015ringit}, in which $T = N, k = 1$} 
\begin{equation}
v_{2k}[n] = \cos(2 \pi n / T), v_{2k+1}[n] = \sin(2 \pi n / T)
\end{equation}
Therefore, the smallest two numerical nonzero eigenvalues each correspond to sinusoids with period $T$ which are mutually orthogonal.  When plotted against each other, they form a circle with an arbitrary phase offset.  Therefore, we compute the circular phase numerically $\phi = \tan^{-1}(\hat{v_1}[n] / \hat{v_2}[n])$, where $\hat{v_i}$ is the eigenvector corresponding to the $i^\text{th}$ smallest nonzero numerical eigenvalue.

@@

\subsection{Persistent Homology}
Our discussion so far has assumed we can find the appropriate $\sigma$ for the (un)weighted graph Laplacian, but this is data dependent.
%The authors of \cite{averbuch2015ringit} use $k$-nearest neighbors, which helps with scale, but which still requires an arbitrary parameter $k$.
To adapt to the data, we leverage 1D persistent homology from topological data analysis \cite{edelsbrunner2010computational}.  This framework allows one to find the scale at which topological features exist in point cloud data.  Specifically, we compute the 1D Vietoris Rips Filtration on our sliding window data $X$, which tracks equivalence classes of loops, known as {\em homology classes} \cite{Hatcher}, as edges and triangles are gradually added between sliding window points.  The algorithm returns a so-called persistence diagram, which is a multiset of points on a 2D birth/death grid, with each point corresponding to a loop class.  The birth value $b_i$ indicates the scale at which the $i^{\text{th}}$ loop class forms, and the death value $d_i$ indicates the scale at which that class no longer exists (expressible as a formal sum of triangle boundaries with coefficients in some field).  The difference $d_i - b_i$ is known as the {\em persistence} of the $i^{\text{th}}$ class.  For sliding window periodic time series, one point in the diagram should have a much larger persistence than the others (e.g. Figure~\ref{fig:ConceptFigure})~\cite{perea2015sliding,tralie2017quasi}, and this ideally reflects the single cycle of motion that we seek.  We take our scale to be $\alpha b_i + (1-\alpha)d_i$ for the largest $d_i - b_i$, where $\alpha \in [0, 1]$ is a parameter which will be explored experimentally in Section~\ref{sec:experiments}.

@@

Given the phase estimates $\phi_i$, we can now synthesize the final slow motion template by lining up the $N$ sliding windows $SW_{d}[\mb{X}(i)]$ by $\phi$.  For a template with $M$ frames, we choose a set of $M$ equally spaced angles $\theta_1, \theta_2, ..., \theta_M$ around the circle at which to sample the template.  Let $\Phi_i$ be the unwrapped phase of $\phi_i$.  Based on this, we esimate the number of periods $X(t)$ goes through as $k = (N \times \max_i(\Phi_i) / (2 \pi (N-d+1)))$.  Assuming $k$ is correct and that $\phi_i$ has a constant slope over $i$, there is a phase gap of $2 \pi k/N$ between adjacent stacked frames in each sliding window.  Thus, $SW_{d}[\mb{X}(i)]$ spans the interval $[\phi_i, \phi_i + 2\pi k d/N (\text{mod} 2 \pi)]$.  A given $\theta_j$ will potentially be contained in the intervals for many different windows, though it is unlikely to coincide exactly with any of the frames in each window, so we use linear interpolation to fill in a frame corresponding to $\theta_j$ in each window overlapping it.  The frame at angle $\phi_j$ is taken as the median of all interpolated frames.  In practice, for videos which go through many periods, the gap between stacked frames within a window may be large, leading to ghosting artifacts in linear interpolation \cite{meyer2015phase}.  We thus add an additional user-defined parameter ``F'' which only includes linearly interpolated frames in the median if the frame to interpolate is at most $F$ frames away from an existing frame in the window. % MSB: I do not understand this last sentence..

@@

We now show some experimental results from our pipeline.  We use the GUDHI library\cite{maria2014gudhi} to compute persistent homology quickly with fields other than $\mathbb{Z}_2$.

>>>>

Our approach to slow motion templates is {\em Eulerian}; that is, we process the video pixel by pixel with no tracking.  Eulerian approaches for video synthesis are attractive due to their simplicity and ease of implementation.  For instance, a Fourier bandpass filtering can successfully elucidate subtle periodic motions in videos \cite{wu2012eulerian, wadhwa2013phase}.  Additionally, time-domain Eulerian approaches have been used to synthesize infinitely playing ``video textures''~\cite{schodl2000video}, as well as looping video templates for more general scenarios~\cite{Liao2013VideoLoops,Liao2015VideoLoops}.  There are some pitfalls with Eulerian techniques, however.  They are known to fail for large motions, and also require the user to manually specify frequencies of interest \cite{wu2012eulerian, wadhwa2013phase}.  Furthermore, period-to-period drift that is due to a moving camera or small variation in motion can be problematic, as well as unrelated motion in the background \cite{stauffer1999adaptive}.

@@

% MSB: opportunity to slim paper by cutting examples 
To address the problems with an Eulerian representation, we devise a geometric and topological Eulerian framework by constructing {\em sliding window embeddings} of videos.  Sliding window embeddings, or delay reconstructions, have found a diverse array of applications in activity recognition \cite{frank2010activity,venkataraman2016shape}, video analysis \cite{tralie2017quasi}, and motion capture analysis \cite{venkataraman2016shape}.  Sliding window embeddings of periodic time series, in particular, form samples of a topological loop \cite{perea2015sliding}, as do generalized multivariate sliding windows on periodic video data, regardless of the type of motion present \cite{traliehigh, tralie2017quasi}.  Furthermore, the sliding window embedding of an Eulerian video provides a form of regularization that mitigates the effects of drift~\cite{tralie2017quasi}.  Our geometric framework also allows us to use topological data analysis (TDA) \cite{edelsbrunner2000topological,edelsbrunner2008persistent,edelsbrunner2010computational,carlsson2009topology,ghrist2014elementary} and fundamental frequency estimation \cite{Mcleod05asmarter} to autotune the spatial and temporal scales of the periodic motion, respectively, so that no user intervention is required.  We use Laplacian Eigenmaps~\cite{belkin2003laplacian} computed on the sliding window embedding to obtain a circular phase $\phi \in [0, 2\pi]$ for each window.  We then reorder the embedding using $\phi$, and reorder the video by having windows that overlap in time vote on final pixels in each frame of the template, further mitigating drift.  Figure~\ref{fig:ConceptFigure} highlights the steps of our technique for a sampled 1D periodic time series.  There are only 12 samples per period in the original signal, so the details of each period are coarse and noisy.  However, once resorted, we get a fine-detailed representation of one period.

% MSB: opportunity to slim paper by cutting examples 
Many prior works use a combination of sliding window embeddings and topological data analysis, for instance detecting chatter in mechanical systems \cite{khasawneh2016chatter} and quantifying periodic activites in motion capture data \cite{vejdemo2015cohomological, venkataraman2016persistent}. However, our approach in using a sliding window jointly with TDA to {\em synthesize} a new result is novel. Related works have used the graph Laplacian to rearrange images around a loop as a pre-processing step for structure from motion \cite{averbuch2015ringit}, and to reorder the frames of microscope images from a developing embryo~\cite{dsilva2015diffusionvecordering}.  However, our use of the graph Laplacian to instead {\em parameterize} time series is unique. One can also view the combination of TDA and Laplacian eigenmaps as a simpler alternative to cohomology circular coordinates \cite{de2011persistent,vejdemo2015cohomological} for the special case of periodic data.
%Our approach is also the opposite of the approach in \cite{bendich2011improving}, which performs diffusion maps before TDA. - MSB: this is related work, not opposite work :)

@@

Figure~\ref{fig:ConceptFigure} shows the different parts of our method for a 1D periodic signal for illustration purposes. In general, the actual input to our method is a 3-channel video of $N$ frames, each frame of resolution $W \times H$. We treat each frame as a point in Euclidean space of dimension $\mathbb{R}^{3 \cdot W \cdot H}$, and denote the set of all videos by the sequence $\mb{X}(t) \in \mathbb{R}^{(3 \cdot W \cdot H)}$ for integer-valued $t \in [1,N]$. The steps of our approach outlined in Figure~\ref{fig:ConceptFigure}, which we split into two parts: an analysis part (a), which culminates in phase estimates $\phi$, and a synthesis part (b), in which the template is created based on information in (a).  For the analysis part, we do not rely on the {\em extrinsic} geometry of the set of points $\mb{X}(t)$, only its set of {\em pairwise distances}. Thus, we can project all of the points into a lower-dimensional space and apply the set of techniques to the projection, assuming that pairwise distances are preserved. Since $N \ll (3 \cdot W \cdot H)$, a distance-preserving projection into an $N-1$-dimensional space is always possible. Denoting this projection by $\mb{P} \in \mathbb{R}^{(N-1) \times (3 \cdot W \cdot H)}$, our method for the steps in (a) operates on $\mb{Z}(t) = \mb{P} \mb{X}(t)$.  We also use the third layer of a Gaussian pyramid on each frame in place of the raw frames before applying $\mb{P}$ to mitigate drift.

@@

The sliding window video embedding \cite{cao1998dynamics,traliehigh,tralie2017quasi} for $\mb{Z}$ is defined as
\begin{equation}
	SW_{d}[\mb{Z}(t)] = \left[ \begin{array}{c} \mb{Z}(t) \\ \mb{Z}(t + 1) \\ \vdots \\ \mb{Z}(t + (d-1))  \end{array} \right] \in \mathbb{R}^{(N-1)\cdot(d)},
\end{equation}
where $d$ is the dimension of the embedding. As shown by Takens~\cite{takens1981detecting}, a sliding window of dimension $2m+1$ of even a single generic observation function of a dynamical system of intrinsic dimension $m$ is sufficient to reconstruct a topological embedding of the underlying trajectory in the original state space.  For periodic signals, this state space is a torus, and $d$ should be twice the number of harmonics present to injectively reconstruct loops on that torus \cite{perea2015sliding}.  The same is true of videos in general \cite{tralie2017quasi}.  Furthermore, the sliding window length $(d-1)$ maximizes the roundness of the embeddings when $(d-1) = k T$ for some integer $k$, where $T$ is the fundamental period.  Since we will use topological data analysis (TDA) to find the scale of the loop, and since TDA measures roundness, it is important that the window size is an integer multiple of the period.  To do this, we perform 1D ISOMAP \cite{tenenbaum2000global} on $\mb{Z}$ to generate a 1D surrogate signal, and perform an autocorrelation-based fundamental frequency estimation \cite{Mcleod05asmarter} on this signal to estimate the period and, hence, the appropriate $d$.  This estimation often returns integer multiples of the fundamental period, which is fine for our purposes.
% MSB: "roundness" is not being very precise

@@

We now examine the unweighted Laplacian of an ideal model to motivate our approach.  Suppose we have extracted a sliding window embedding $SW_{d}$ for a video $\mb{Z}$ which repeats itself exactly every $T$ frames for $k$ times, for a total of $N = kT$ windows.  Let us further suppose it is possible to choose an $\sigma$ so that the unweighted adjacency matrix contains an entry of 1 for edges corresponding to windows that are adjacent in time, as well as for the corresponding windows at, before, and after repetitions of the window at intervals of $T$; that is, the Laplacian is a symmetric, {\em circulant} matrix defined as

@@

The smallest two nonzero eigenvalues occur when $m = k$, corresponding to the eigenvectors \footnote{This generalizes the circle graph used in \cite{averbuch2015ringit}, in which $T = N, k = 1$} 
\begin{equation}
v_{2k}[n] = \cos(2 \pi n / T), v_{2k+1}[n] = \sin(2 \pi n / T)
\end{equation}
Therefore, the smallest two numerically nonzero eigenvalues each correspond to sinusoids with period $T$ which are mutually orthogonal.  When plotted against each other, they form a circle with an arbitrary phase offset.  Therefore, we compute the circular phase numerically $\phi = \tan^{-1}(\hat{v_1}[n] / \hat{v_2}[n])$, where $\hat{v_i}$ is the eigenvector corresponding to the $i^\text{th}$ smallest numerically nonzero eigenvalue.

@@

\subsection{Persistent Homology}
Our discussion so far has assumed we can find the appropriate $\sigma$ for the (un)weighted graph Laplacian, but this is data dependent.
%The authors of \cite{averbuch2015ringit} use $k$-nearest neighbors, which helps with scale, but which still requires an arbitrary parameter $k$.
To adapt to the data, we leverage 1D persistent homology from topological data analysis \cite{edelsbrunner2010computational}.  This framework allows one to find the scale at which topological features exist in point cloud data.  Specifically, we compute the 1D Vietoris Rips Filtration on our sliding window data $X$, which tracks equivalence classes of loops, known as {\em homology classes} \cite{Hatcher}, as edges and triangles are gradually added between sliding window points.  The algorithm returns a so-called persistence diagram, which is a multiset of points on a 2D birth/death grid, with each point corresponding to a loop class.  The birth value $b_i$ indicates the scale at which the $i^{\text{th}}$ loop class forms, and the death value $d_i$ indicates the scale at which that class no longer exists.  The difference $d_i - b_i$ is known as the {\em persistence} of the $i^{\text{th}}$ class.  For sliding window periodic time series, one point in the diagram should have a much larger persistence than the others (e.g. Figure~\ref{fig:ConceptFigure})~\cite{perea2015sliding,tralie2017quasi}, and this ideally reflects the single cycle of motion that we seek.  We take our scale to be $\alpha b_i + (1-\alpha)d_i$ for the largest $d_i - b_i$, where $\alpha \in [0, 1]$ is a parameter which will be explored experimentally in Section~\ref{sec:experiments}.

@@

Given the phase estimates $\phi_i$, we can now synthesize the final slow motion template by lining up the $N$ sliding windows $SW_{d}[\mb{X}(i)]$ by $\phi$.  For a template with $M$ frames, we choose a set of $M$ equally spaced angles $\theta_1, \theta_2, ..., \theta_M$ around the circle at which to sample the template.  Let $\Phi_i$ be the unwrapped phase of $\phi_i$.  Based on this, we esimate the number of periods $\mb{X}$ goes through as $k = (N \times \max_i(\Phi_i) / (2 \pi (N-d+1)))$.  Assuming $k$ is correct and that $\phi_i$ has a constant slope over $i$, there is a phase gap of $2 \pi k/N$ between adjacent stacked frames in each sliding window.  Thus, $SW_{d}[\mb{X}(i)]$ spans the interval $[\phi_i, \phi_i + 2\pi k d/N (\text{mod} 2 \pi)]$.  A given $\theta_j$ will potentially be contained in the intervals for many different windows, though it is unlikely to coincide exactly with any of the frames in each window, so we use linear interpolation to fill in a frame corresponding to $\theta_j$ in each window overlapping it.  The frame at angle $\phi_j$ is taken as the median of all interpolated frames.  In practice, for videos which go through many periods, the gap between stacked frames within a window may be large, leading to ghosting artifacts in linear interpolation \cite{meyer2015phase}.  We thus add an additional user-defined parameter ``F'' which only includes linearly interpolated frames in the median if the frame to interpolate is at most $F$ frames away from an existing frame in the window. % MSB: I do not understand this last sentence, but could probably be tighter..

@@

We now show some experimental results from our pipeline.
%We use the GUDHI library\cite{maria2014gudhi} to compute persistent homology quickly with fields other than $\mathbb{Z}_2$.
